Limited plasticity but increased variance in physiological rates across ectotherm populations under climate change

### 2024-07-14 ms.qmd
Fix line ending drivig me nuts and rerender

Climate change is expected to result in warmer and more variable thermal
environments globally. Maintaining variability in physiological rates
and adjusting them in response to extreme temperatures (plasticity) is
essential for allowing populations to adapt to climate change. Yet,
incorporating both plasticity and changes in phenotypic variation when
predicting the impacts of climate change on populations is often
ignored. We developed effect size estimates that quantify how
physiological rate variance changes relative to the mean as temperature
increases, which we refer to as 'relative variance'. We show by
meta-analysing \>1900 effects from 226 species that relative variance in
physiological rates decreases at higher temperatures driven by the lower
proportional increase in physiological rate variance relative to the
mean. Freshwater ectotherms are capable of acclimating and have the
smallest reductions in relative variance. Marine organisms also
acclimated, but plasticity traded-off with a reduction in relative
variance in physiological rates at higher temperatures. Relative
variance reductions were particularly pronounced for terrestrial
ectotherms, and this coincided with a lack of capacity for acclimation,
highlighting the vulnerability of terrestrial ectotherms to climate
change. Our results show that, while there are weak acclimation
responses in some habitats (marine and freshwater), such responses may
trade-off with reductions in relative physiological rate variance that
may have important evolutionary and ecological ramifications that affect
our understanding of how climate change will impact populations now and
in the future.

@@

We analysed our data using multilevel meta-analytic (MLMA) and
meta-regression (MLMR) models in R (vers.
`r paste0(R.Version()$major, ".", R.Version()$minor)`) using *brms*
[vers. `r utils::packageVersion("brms")` @Burkner2017-wg;
@Burkner2018-ur; @noauthor_2021-az] and *metafor* [vers.
`r utils::packageVersion("metafor")` @Viechtbauer2010-fn]. We fit both
Bayesian and frequentist approaches to ensure that our results were
consistent, and to create orchard plots more easily [vers. 2.0,
@Nakagawa2021-ls; @nakagawaorchard]. In addition, Bayesian methods
better protect against type I errors in the presence of complex sources
of non-independence [@Nakagawa2021-uf; @Song2021-ba; @Noble2017-mz]. In
all cases, frequentist and Bayesian models resulted in the same
conclusions. For our Bayesian models, we ran 4 MCMC chains, each with a
warm-up of 1000 followed by 4000 sampling iterations keeping every 5
iterations for a minimum of 3200 samples from the posterior
distribution. We used flat Gaussian priors for 'fixed' effects (i.e.,
$N(0, 10)$) and a student t-distribution for 'random' effects (i.e.,
$student_{t}(3, 0, 10)$). We checked that all MCMC chains were mixing
and had converged (i.e., $R_{hat} = 1$). We compared any competing
models using Akaike's Information Criteria (AIC) (if frequentist) or
Wantabe Information Criteria (WAIC) (if Bayesian). We deemed models with
the lowest IC value to be best supported if there was a $\Delta IC$
between the competing models of 2 or more. If two models were within 2
$\Delta IC$ units we went with the most parsimonious model.

@@

After quantifying levels of heterogeneity, we fit a series of
multi-level meta-regression (MLMR) models to test our key questions. In
all models, we included the same random effects as we used in our MLMA
models. Acclimation time varied from
`r range(data_wide$acclim_period, na.rm = TRUE)[1]` to
`r range(data_wide$acclim_period, na.rm = TRUE)[2]` days (mean $\pm$ SD
= `r mean(data_wide$acclim_period, na.rm = TRUE)` $\pm$
`r sd(data_wide$acclim_period, na.rm = TRUE)` days), and terrestrial
ectotherms were acclimated for a much shorter duration (mean $\pm$ SD =
`r acclim_time[3,2]` $\pm$ `r acclim_time[3,3]` days, n =
`r acclim_time[3,4]`) than both freshwater (mean $\pm$ SD =
`r acclim_time[1,2]` $\pm$ `r acclim_time[1,3]` days, n =
`r acclim_time[1,4]`) and marine species (mean $\pm$ SD =
`r acclim_time[2,2]` $\pm$ `r acclim_time[2,3]` days, n =
`r acclim_time[2,4]`). However, rates of acclimation have been shown to
be faster for many terrestrial groups compared to aquatic organisms
\[e.g., amphibians and reptiles have faster rates of acclimation than
fishes; See @Einum2023-lz\]. Nonetheless, to control for these possible
differences, acclimation time was mean-centered (mean = 0) and included
in all our models, though it was not a strong predictor of effect size
variation (*Supplementary Materials*, @fig-accl_time).

@@

Second, we tested whether acute and acclimation $lnRR_{Q_{10}}$ and
$lnVR_{Q_{10}}$ differed between whole-organism versus biochemical
traits across habitats by fitting an model with an interaction between
type, habitat and trait category (referred to as "Model 3"). We also
explored how acute and acclimation effects varied across more detailed
trait categories, and these results can be found in the *Supplementary
Materials*. Given that residual variance appeared to differ across trait
categories we fit heteroscedastic residual variance models which were
better supported in all cases ($\Delta AIC_{c}$; marine = 58, freshwater
= 120, and terrestrial = 12). We expected that whole-organism traits
would be more likely to maintain relative variation in physiological
function and be less likely to acclimate because whole-organism function
relies on a greater number of biochemical reactions each with different
thermal sensitivies [@fields2001protein; @angilletta2009thermal].

Third, we tested whether different life-stages were more or less likely
to acclimate by fitting a model for each habitat type and including an
interaction between life-stage ('adult' or 'juvenile') and effect type
(referred to as "Model 4"). We predicted that acclimation responses
would be more likely early in development compared to later in
development as this pattern has been shown in previous studies [e.g.,
@moghadam2019heat], but that this should depend on the habitat type
given the different constraints faced by different early life stages
across major habitat types.

@@

### *Performance Curve Simulation*
To better understand the characteristics of the performance curves in a sample that would lead to observed changes in variance across temperature we conducted a simple simulation. To simulate performance curves, we used an asymmetrical Gaussian function:

@@

The final dataset included a total of
`r f <- taxa_sum |> dplyr::filter(habitat == "f") |> dplyr::summarise(spp = sum(spp), k = sum(n)); f[1,1]`
freshwater (fishes =
`r taxa_sum |> dplyr::filter(habitat == "f" & class == "Actinopterygii") |> dplyr::summarise(sp = sum(spp))`,
molluscs =
`r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`,
amphibians =
`r taxa_sum |> dplyr::filter(habitat == "f" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))`,
reptiles =
`r taxa_sum |> dplyr::filter(habitat == "f" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`,
arthropods =
`r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`,
and a single crustacean and nematode species),
`r m <- taxa_sum |> dplyr::filter(habitat == "m") |> dplyr::summarise(spp = sum(spp), k = sum(n)); m[1,1]`
marine (fishes =
`r taxa_sum |> dplyr::filter(habitat == "m" & class %in% c("Actinopterygii", "Agnatha","Chondrichthyes")) |> dplyr::summarise(sp = sum(spp))`,
annelids =
`r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`,
molluscs =
`r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`,
echinoderms =
`r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Echinodermata") |> dplyr::summarise(sp = sum(spp))`,
reptiles =
`r taxa_sum |> dplyr::filter(habitat == "m" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`,
arthropods =
`r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`,
and a single crustacean and cnidarian species), and
`r t <- taxa_sum |> dplyr::filter(habitat == "t") |> dplyr::summarise(spp = sum(spp), k = sum(n)); t[1,1]`
terrestrial species (annelids =
`r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`,
molluscs =
`r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`,
arthropods =
`r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`,
reptiles =
`r taxa_sum |> dplyr::filter(habitat == "t" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`
and amphibians =
`r taxa_sum |> dplyr::filter(habitat == "t" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))`
along with a single tardigrade species) (@fig-phylo). We had more data
on acute thermal responses (n =
`r trimmed <- data_long2 |> dplyr::filter(grepl("lnRR", name) & !is.na(effect_size) & !is.na(V) & !V == 0); sum1 = trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acute"); sum(sum1$n)`)
compared to thermal responses after an acclimation period (n =
`r sum2 <- trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acclim"); sum2$n`)
because acute responses were reported for each of the two acclimation
temperatures (@fig-phylo). The two acute $lnRR_{Q_{10}}$ effect sizes
([@fig-concept]A) differed significantly from each other ($\beta$ =
`r mean(change_acute)`, 95% CI: `r quantile(change_acute, 0.025)` to
`r quantile(change_acute, 0.975)`, $p_{MCMC}$ =
`r p_value(pmcmc(change_acute))`) with animals acclimated to high
temperatures having slightly higher average $lnRR_{Q_{10}}$ ($\mu$ =
`r mean(hot_acute)`, 95% CI: 95% CI: `r quantile(hot_acute, 0.025)` to
`r quantile(hot_acute, 0.975)`, $p_{MCMC}$ =
`r p_value(pmcmc(hot_acute))`, $Q_{10}$ = `r mean(exp(hot_acute))`)
compared to animals at lower temperatures ($\mu$ = `r mean(cold_acute)`,
95% CI: 95% CI: `r quantile(cold_acute, 0.025)` to
`r quantile(cold_acute, 0.975)`, $p_{MCMC}$ =
`r p_value(pmcmc(cold_acute))`, $Q_{10}$ = `r mean(exp(cold_acute))`)
(@fig-Sacute). However, on average they were in the same direction and
only differed by \~10%. As such, we averaged the two acute
$lnRR_{Q_{10}}$ effect sizes in subsequent analyses.

@@

Effect heterogeneity was high (only `r mean(tot_i2)`% of the variance
was the result of sampling variability, 95% CI:
`r quantile(tot_i2, 0.025)[1]` to `r quantile(tot_i2, 0.975)[1]`%), and
most variance was explained by the specific study and type of trait
(Study: `r mean(stdy_i2)`% , 95% CI: `r quantile(stdy_i2, 0.025)[1]` to
`r quantile(stdy_i2, 0.975)[1]`%; Trait Type: `r mean(trait_i2)`% , 95%
CI: `r quantile(trait_i2, 0.025)[1]` to
`r quantile(trait_i2, 0.975)[1]`%). Evolutionary relationships among
taxa and species ecology (i.e., species random effect) explained little
variation in acute and acclimation responses (Species: `r mean(spp_i2)`%
, 95% CI: `r quantile(spp_i2, 0.025)[1]` to
`r quantile(spp_i2, 0.975)[1]`%; Phylogeny: `r mean(phylo_i2)`% , 95%
CI: `r quantile(phylo_i2, 0.025)[1]` to
`r quantile(phylo_i2, 0.975)[1]`%). All results are from Model 1 (see
“Meta-Analysis” above).

@@

```{r, fig-2}
#| label: fig-2
#| fig-cap: "(A) Mean acute and acclimation $lnRR_{Q_{10}}$ across ectotherms in marine, freshwater, and terrestrial habitats. (B) Mean acute and acclimation $lnVR_{Q_{10}}$ across ectotherms in marine, freshwater and terrestrial habitats. For both plots, k = total number of effect size estimates while the numbers in brackets indicate the number of species. Sample sizes are the same for panel A and B. In both panels, thick bars are 95% confidence intervals (CI) and thin bars 95% prediction intervals (PI). $\\Beta$ values are the contrasts between acute and acclimation means within each habitat. $\\mu$ values are the overall meta-analytic means averaged across acute and acclimation types within each habitat type. In both cases, their 95% CI's are indicated within square brackets. $p_{MCMC}$ values are the posterior probability of the contrast or overall meta-analytic mean being different from zero. For ease of visualisation, all the raw data plotted for both acute and acclimation type effect sizes are presented as circles."
#| warning: false
#| echo: false

@@

```{r, fig-btrait}
#| label: fig-btrait
#| fig-cap: "Estimated mean acclimation and acute $lnRR_{Q_{10}}$ (A-C) and $lnVR_{Q_{10}}$ (D-F) effect sizes for tissue/whole-orgamism traits and biochemical traits across marine (A & D), freshwater (B & E) and terrestrial (C & F) habitats. Across all plots, thick bars indicate 95% confidence intervals and thin bars indicate 95% prediction intervals. k = total number of effect size estimates while the numbers in brackets indicate the number of species. For ease of visualisation, raw data for both trait categories are presented but points are not distinguished by different symbols. "
#| echo: false
#| warning: false

@@

```{r, fig-cvrlh}
#| label: fig-cvrlh
#| fig-cap: "Estimated mean acclimation and acute $lnRR_{Q_{10}}$ (A-C) and $lnVR_{Q_{10}}$ (D-F) for adult (a) and juvenile (j) life-history stages for marine (A & D), freshwater (B & E) and terrestrial (C & F) ectotherms. Across all plots, thick bars indicate 95% confidence intervals and thin bars indicate 95% prediction intervals. k = total number of effect size estimates while the numbers in brackets indicate the number of species. For ease of visualisation, raw data for both adult and juvenile life-history stages are presented but points are not distinguished by different symbols. $\\Beta$ values are the contrasts between acute and acclimation means within each life stage. $p_{MCMC}$ values are the posterior probability of the contrast being different from zero."
#| echo: false
#| warning: false

@@

Variance in physiological rates ($lnVR_{Q_{10}}$) showed an
increase with increasing temperature across all habitat types ([@fig-2]B) with less heterogeneity overall ($I^2_{sv}$ = `r mean(tot_i2_VR)`, 95% CI:
`r quantile(tot_i2_VR, 0.025)` to `r quantile(tot_i2_VR, 0.975)`; "Model 1"; see *Supplementary Materials* @fig-s3). However, physiological rate variance increased significantly more in freshwater for acute resposnes (freshwater vs terrestrial: $\beta$ = `r mean(f_m_c_a)`, 95% CI: `r quantile(f_m_c_a, 0.025)` to `r quantile(f_m_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_a))`), but not for acclimation responses because increases in rates were dampended by acclimation resulting in smaller increases in variance (freshwater vs terrestrial: $\beta$ = `r mean(f_m_c_aclim)`, 95% CI: `r quantile(f_m_c_aclim, 0.025)` to `r quantile(f_m_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_aclim))`). While marine ectotherms had larger increases in physiological rate variance compared to terrestrial ectoterms these were not signigicant (Acute: $\beta$ = `r mean(m_t_c_a)`, 95% CI: `r quantile(m_t_c_a, 0.025)` to `r quantile(m_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_a))`; Acclimation: $\beta$ = `r mean(m_t_c_aclim)`, 95% CI: `r quantile(m_t_c_aclim, 0.025)` to `r quantile(m_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_aclim))`)([@fig-2]B). Marine and freshwater did not differ in how much variance changed at higher temperatures (Acute: $\beta$ = `r mean(f_m_c_a)`, 95% CI: `r quantile(f_m_c_a, 0.025)` to `r quantile(f_m_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_a))`; Acclimation: $\beta$ = `r mean(f_m_c_aclim)`, 95% CI: `r quantile(f_m_c_aclim, 0.025)` to `r quantile(f_m_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_aclim))`). There were no differences between
$lnVR_{Q_{10_{acute}}}$ and $lnVR_{Q_{10_{acclimation}}}$ in any habitat ([@fig-2]B). Differences
in variance between habitats were driven by different scaling relationships between mean and variance with slopes varying across habitat types (see *Supplementary
Materials* analyses with $lnCVR_{Q_{10}}$; @tbl-slopes & @fig-mv).

>>>>

Climate change is expected to result in warmer and more variable thermal environments globally. Maintaining variability in physiological rates and adjusting them in response to extreme temperatures (plasticity) is essential for allowing populations to adapt to climate change. Yet, incorporating both plasticity and changes in phenotypic variation when predicting the impacts of climate change on populations is often ignored. We developed effect size estimates that quantify how physiological rate variance changes relative to the mean as temperature increases, which we refer to as 'relative variance'. We show by meta-analysing >1900 effects from 226 species that relative variance in physiological rates decreases at higher temperatures driven by the lower proportional increase in physiological rate variance relative to the mean. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also acclimated, but plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Our results show that, while there are weak acclimation responses in some habitats (marine and freshwater), such responses may trade-off with reductions in relative physiological rate variance that may have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations now and in the future.

@@

We analysed our data using multilevel meta-analytic (MLMA) and meta-regression (MLMR) models in R (vers.
`r paste0(R.Version()$major, ".", R.Version()$minor)`) using *brms* [vers. `r utils::packageVersion("brms")` @Burkner2017-wg;
@Burkner2018-ur; @noauthor_2021-az] and *metafor* [vers. `r utils::packageVersion("metafor")` @Viechtbauer2010-fn]. We fit both Bayesian and frequentist approaches to ensure that our results were consistent, and to create orchard plots more easily [vers. 2.0,
@Nakagawa2021-ls; @nakagawaorchard]. In addition, Bayesian methods better protect against type I errors in the presence of complex sources of non-independence [@Nakagawa2021-uf; @Song2021-ba; @Noble2017-mz]. In all cases, frequentist and Bayesian models resulted in the same conclusions. For our Bayesian models, we ran 4 MCMC chains, each with a warm-up of 1000 followed by 4000 sampling iterations keeping every 5 iterations for a minimum of 3200 samples from the posterior distribution. We used flat Gaussian priors for 'fixed' effects (i.e.,$N(0, 10)$) and a student t-distribution for 'random' effects (i.e., $student_{t}(3, 0, 10)$). We checked that all MCMC chains were mixing and had converged (i.e., $R_{hat} = 1$). We compared any competing models using Akaike's Information Criteria (AIC) (if frequentist) or Wantabe Information Criteria (WAIC) (if Bayesian). We deemed models with the lowest IC value to be best supported if there was a $\Delta IC$ between the competing models of 2 or more. If two models were within 2 $\Delta IC$ units we went with the most parsimonious model.

@@

After quantifying levels of heterogeneity, we fit a series of multi-level meta-regression (MLMR) models to test our key questions. In
all models, we included the same random effects as we used in our MLMA models. Acclimation time varied from `r range(data_wide$acclim_period, na.rm = TRUE)[1]` to `r range(data_wide$acclim_period, na.rm = TRUE)[2]` days (mean $\pm$ SD
= `r mean(data_wide$acclim_period, na.rm = TRUE)` $\pm$ `r sd(data_wide$acclim_period, na.rm = TRUE)` days), and terrestrial
ectotherms were acclimated for a much shorter duration (mean $\pm$ SD = `r acclim_time[3,2]` $\pm$ `r acclim_time[3,3]` days, n =
`r acclim_time[3,4]`) than both freshwater (mean $\pm$ SD = `r acclim_time[1,2]` $\pm$ `r acclim_time[1,3]` days, n =
`r acclim_time[1,4]`) and marine species (mean $\pm$ SD = `r acclim_time[2,2]` $\pm$ `r acclim_time[2,3]` days, n =
`r acclim_time[2,4]`). However, rates of acclimation have been shown to be faster for many terrestrial groups compared to aquatic organisms [e.g., amphibians and reptiles have faster rates of acclimation than fishes; See @Einum2023-lz]. Nonetheless, to control for these possible differences, acclimation time was mean-centered (mean = 0) and included in all our models, though it was not a strong predictor of effect size variation (*Supplementary Materials*, @fig-accl_time).

@@

Second, we tested whether acute and acclimation $lnRR_{Q_{10}}$ and $lnVR_{Q_{10}}$ differed between whole-organism versus biochemical traits across habitats by fitting an model with an interaction between type, habitat and trait category (referred to as "Model 3"). We also explored how acute and acclimation effects varied across more detailed trait categories, and these results can be found in the *Supplementary Materials*. Given that residual variance appeared to differ across trait categories we fit heteroscedastic residual variance models. We expected that whole-organism traits would be more likely to maintain relative variation in physiological function and be less likely to acclimate because whole-organism function relies on a greater number of biochemical reactions each with different thermal sensitivies [@fields2001protein; @angilletta2009thermal].

Third, we tested whether different life-stages were more or less likely to acclimate by fitting a model for each habitat type and including an interaction between life-stage ('adult' or 'juvenile') and effect type (referred to as "Model 4"). We predicted that acclimation responses would be more likely early in development compared to later in development as this pattern has been shown in previous studies [e.g.,@moghadam2019heat], but that this should depend on the habitat type given the different constraints faced by different early life stages across major habitat types.

@@

### *Simulating among-individual variance in performance curves*
To better understand the characteristics of the performance curves in a sample that would lead to observed changes in variance across temperature we conducted a simple simulation. To simulate performance curves, we used an asymmetrical Gaussian function:

@@

The final dataset included a total of `r f <- taxa_sum |> dplyr::filter(habitat == "f") |> dplyr::summarise(spp = sum(spp), k = sum(n)); f[1,1]` freshwater (fishes = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Actinopterygii") |> dplyr::summarise(sp = sum(spp))`, molluscs = `r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`,
amphibians = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))`,
reptiles = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`,
arthropods = `r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`,
and a single crustacean and nematode species), `r m <- taxa_sum |> dplyr::filter(habitat == "m") |> dplyr::summarise(spp = sum(spp), k = sum(n)); m[1,1]` marine (fishes = `r taxa_sum |> dplyr::filter(habitat == "m" & class %in% c("Actinopterygii", "Agnatha","Chondrichthyes")) |> dplyr::summarise(sp = sum(spp))`, annelids = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`, molluscs = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, echinoderms = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Echinodermata") |> dplyr::summarise(sp = sum(spp))`, reptiles = `r taxa_sum |> dplyr::filter(habitat == "m" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`, arthropods = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`,
and a single crustacean and cnidarian species), and `r t <- taxa_sum |> dplyr::filter(habitat == "t") |> dplyr::summarise(spp = sum(spp), k = sum(n)); t[1,1]` terrestrial species (annelids = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`, molluscs = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, arthropods = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`,
reptiles = `r taxa_sum |> dplyr::filter(habitat == "t" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))` and amphibians =
`r taxa_sum |> dplyr::filter(habitat == "t" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))` along with a single tardigrade species) (@fig-phylo). We had more data on acute thermal responses (n = `r trimmed <- data_long2 |> dplyr::filter(grepl("lnRR", name) & !is.na(effect_size) & !is.na(V) & !V == 0); sum1 = trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acute"); sum(sum1$n)`) compared to thermal responses after an acclimation period (n = `r sum2 <- trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acclim"); sum2$n`) because acute responses were reported for each of the two acclimation temperatures (@fig-phylo). The two acute $lnRR_{Q_{10}}$ effect sizes ([@fig-concept]A) differed significantly from each other ($\beta$ = `r mean(change_acute)`, 95% CI: `r quantile(change_acute, 0.025)` to `r quantile(change_acute, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(change_acute))`) with animals acclimated to high temperatures having slightly higher average $lnRR_{Q_{10}}$ ($\mu$ = `r mean(hot_acute)`, 95% CI: 95% CI: `r quantile(hot_acute, 0.025)` to
`r quantile(hot_acute, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(hot_acute))`, $Q_{10}$ = `r mean(exp(hot_acute))`) compared to animals at lower temperatures ($\mu$ = `r mean(cold_acute)`, 95% CI: 95% CI: `r quantile(cold_acute, 0.025)` to `r quantile(cold_acute, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(cold_acute))`, $Q_{10}$ = `r mean(exp(cold_acute))`) (@fig-Sacute). However, on average they were in the same direction and only differed by ~10%. As such, we averaged the two acute $lnRR_{Q_{10}}$ effect sizes in subsequent analyses.

@@

Effect heterogeneity was high (only `r mean(tot_i2)`% of the variance was the result of sampling variability, 95% CI: `r quantile(tot_i2, 0.025)[1]` to `r quantile(tot_i2, 0.975)[1]`%), and most variance was explained by the specific study and type of trait (Study: `r mean(stdy_i2)`% , 95% CI: `r quantile(stdy_i2, 0.025)[1]` to `r quantile(stdy_i2, 0.975)[1]`%; Trait Type: `r mean(trait_i2)`% , 95% CI: `r quantile(trait_i2, 0.025)[1]` to `r quantile(trait_i2, 0.975)[1]`%). Evolutionary relationships among taxa and species ecology (i.e., species random effect) explained little variation in acute and acclimation responses (Species: `r mean(spp_i2)`%, 95% CI: `r quantile(spp_i2, 0.025)[1]` to `r quantile(spp_i2, 0.975)[1]`%; Phylogeny: `r mean(phylo_i2)`% , 95%
CI: `r quantile(phylo_i2, 0.025)[1]` to `r quantile(phylo_i2, 0.975)[1]`%). All results are from Model 1 (see “Meta-Analysis” above).

@@

```{r, fig-2}
#| label: fig-2
#| fig-cap: "(A) Mean acute and acclimation $lnRR_{Q_{10}}$ across ectotherms in marine, freshwater, and terrestrial habitats. (B) Mean acute and acclimation $lnVR_{Q_{10}}$ across ectotherms in marine, freshwater and terrestrial habitats. For both plots, k = total number of effect size estimates while the numbers in brackets indicate the number of species. Sample sizes are the same for panel A and B. In both panels, thick bars are 95% confidence intervals (CI) and thin bars 95% prediction intervals (PI). $\\beta$ values are the contrasts between acute and acclimation means within each habitat. $\\mu$ values are the overall meta-analytic means averaged across acute and acclimation types within each habitat type. In both cases, their 95% CI's are indicated within square brackets. $p_{MCMC}$ values are the posterior probability of the contrast or overall meta-analytic mean being different from zero. For ease of visualisation, all the raw data plotted for both acute and acclimation type effect sizes are presented as circles."
#| warning: false
#| echo: false

@@

```{r, fig-btrait}
#| label: fig-btrait
#| fig-cap: "Estimated mean acclimation and acute $lnRR_{Q_{10}}$ (A-C) and $lnVR_{Q_{10}}$ (D-F) effect sizes for tissue/whole-orgamism traits and biochemical traits across terrestrial (A & D), marine (B & E) and freshwater (C & F) ectotherms. Across all plots, thick bars indicate 95% confidence intervals and thin bars indicate 95% prediction intervals. k = total number of effect size estimates while the numbers in brackets indicate the number of species. For ease of visualisation, raw data for both trait categories are presented but points are not distinguished by different symbols. "
#| echo: false
#| warning: false

@@

```{r, fig-cvrlh}
#| label: fig-cvrlh
#| fig-cap: "Estimated mean acclimation and acute $lnRR_{Q_{10}}$ (A-C) and $lnVR_{Q_{10}}$ (D-F) for adult (a) and juvenile (j) life-history stages for terrestrial (A & D), marine (B & E) and freshwater (C & F) ectotherms. Across all plots, thick bars indicate 95% confidence intervals and thin bars indicate 95% prediction intervals. k = total number of effect size estimates while the numbers in brackets indicate the number of species. For ease of visualisation, raw data for both adult and juvenile life-history stages are presented but points are not distinguished by different symbols. $\\beta$ values are the contrasts between acute and acclimation means within each life stage. $p_{MCMC}$ values are the posterior probability of the contrast being different from zero."
#| echo: false
#| warning: false

@@

Variance in physiological rates ($lnVR_{Q_{10}}$) showed an increase with increasing temperature across all habitat types ([@fig-2]B) with less heterogeneity overall ($I^2_{sv}$ = `r mean(tot_i2_VR)`, 95% CI: `r quantile(tot_i2_VR, 0.025)` to `r quantile(tot_i2_VR, 0.975)`; "Model 1"; see *Supplementary Materials* @fig-s3). However, physiological rate variance increased significantly more in freshwater for acute responses (freshwater vs terrestrial: $\beta$ = `r mean(f_t_c_a)`, 95% CI: `r quantile(f_t_c_a, 0.025)` to `r quantile(f_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_t_c_a))`), but not for acclimation responses because increases in rates were dampened by acclimation resulting in smaller increases in variance (freshwater vs terrestrial: $\beta$ = `r mean(f_t_c_aclim)`, 95% CI: `r quantile(f_t_c_aclim, 0.025)` to `r quantile(f_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_t_c_aclim))`). While marine ectotherms had larger increases in physiological rate variance compared to terrestrial ectotherms these were not significant (Acute: $\beta$ = `r mean(m_t_c_a)`, 95% CI: `r quantile(m_t_c_a, 0.025)` to `r quantile(m_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_a))`; Acclimation: $\beta$ = `r mean(m_t_c_aclim)`, 95% CI: `r quantile(m_t_c_aclim, 0.025)` to `r quantile(m_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_aclim))`)([@fig-2]B). Marine and freshwater did not differ in how much variance changed at higher temperatures (Acute: $\beta$ = `r mean(f_m_c_a)`, 95% CI: `r quantile(f_m_c_a, 0.025)` to `r quantile(f_m_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_a))`; Acclimation: $\beta$ = `r mean(f_m_c_aclim)`, 95% CI: `r quantile(f_m_c_aclim, 0.025)` to `r quantile(f_m_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_aclim))`). There were no differences between $lnVR_{Q_{10_{acute}}}$ and $lnVR_{Q_{10_{acclimation}}}$ in any habitat ([@fig-2]B). Differences in variance between habitats were driven by different scaling relationships between mean and variance with slopes varying across habitat types (see *Supplementary Materials* analyses with $lnCVR_{Q_{10}}$; @tbl-slopes & @fig-mv).

Overall, there was a `r mean(t_VR_mean_ef)`% (95% CI: `r quantile(t_VR_mean_ef, 0.025)[1]` to `r quantile(t_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(t_VR_mean))`) increase in physiological rate variance for terrestrial ectotherms and a `r mean(m_VR_mean_ef)`% (95% CI:`r quantile(m_VR_mean_ef, 0.025)[1]` to `r quantile(m_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(m_VR_mean))`) increase in variation for marine ectotherms when temperatures increased by 10$^\circ$C. In contrast, freshwater ectotherms exhibited a `r mean(f_VR_mean_ef)`%, 95% CI: `r quantile(f_VR_mean_ef, 0.025)[1]` to `r quantile(f_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(f_VR_mean))` increase in variance at high temperatures (@fig-2; results from "Model 2"). The relative differences between habitats were similar when considering $lnCVR_{Q_{10}}$ (See *Supplementary Materials*).
